Asymmetric relationships and their impact on coexistence: Supporting information

### 2023-01-26 paper/SupMat.Rnw
edited Theory section and caption of fig 2 and corresponding sections of sup mat

@@

For the lottery model, for $i \neq j$, we have
\begin{align}
\sC_j &= -\ln \left[ 1-\delta+\delta N \frac{\overline{B}_j}{B_1(t) N_1 (t) + B_2(t) N_2(t)}  \right]  \\
\sC_{j \bs i} &= -\ln \left[ 1-\delta+\delta  \frac{\overline{B}_j}{B_j(t)} \right] \\
\sC_{i \bs i} &= -\ln \left[ 1-\delta+\delta  \frac{\overline{B}_i}{B_j(t)} \right].
\end{align}
Both $\sC_{i \bs i}$ and $\sC_{j \bs i}$ are strictly monotonic functions of $B_j$, so the function
$B_j \mapsto \sC_{j \bs i}$ is invertible, and using its inverse provides an expression of $\sC_{i \bs i}$ 
in terms of $\sC_{j \bs i}$. One could write this expression explicitly and then differentiate it to calculate $q_{ij}$,
but we instead compute $q_{ij}$ through implicit differentiation:
\begin{align}
\frac{\partial \sC_{i \bs i}}{\partial \sC_{j \bs i}} &= \left. \frac{\partial \sC_{i \bs i}}{\partial B_j} \middle/ \frac{\partial \sC_{j \bs i}}{\partial B_j} \right. \\
&= \left( \frac{1-\delta +\delta \frac{\overline{B}_j}{B_j}}{1-\delta+\delta \frac{\overline{B}_i}{B_j}} \right) \left( \frac{\overline{B}_i}{\overline{B}_j}  \right).
\end{align}
Solving $\sC_{j \bs i}=0$ gives $B_j=\overline{B}_j$, so 
\begin{align}
q_{ij} &= \left( \frac{1-\delta +\delta \frac{\overline{B}_j}{\overline{B}_j}}{1-\delta+\delta \frac{\overline{B}_i}{\overline{B}_j}} \right) \left( \frac{\overline{B}_i}{\overline{B}_j}  \right) \\
&= \frac{\overline{B}_i}{\overline{B}_j (1-\delta) +\overline{B}_i \delta},
\end{align}
as stated in the main text.

%***DAN: Change the below paragraph if needed. Some reasons it might be needed: Steve joins as a coauthor - easy
%just means a change in wording at the top bit;
%Steve reads this and thinks it makes sense to change it because something is factually or presentationally 
%sub-optimal; or Dan or Jasmin actually develops a more rigorous understanding of scaling factors and based
%on that thinks something here is sub-optimal. Anyway, the "do both" approach seems like it can't go wrong,
%especially if Steve's prediction is correct that it won't make much difference.
%***DAN: Dan, you probably need to spend a bit more time with the cited papers to better understand this. 
We also use $q_{ij}=1$ as an alternative value, for the following reasons.
First, as discussed in the supporting information of \cite{Ellner_2016}, and as further elaborated by 
\cite{Barabas_2018}, it takes very special circumstances for the scaling factors to exist and be unique, 
especially with size-structured models. Second, in empirically parameterized models, scaling factor values are 
often not robust to arbitrary modelling choices. 
Although the lottery model is not size structured and we do not empirically parameterize it, the above shortcomings of 
scaling factors, generally, suggest we should find alternatives to their use. The alternative 
suggested by Ellner for empirical applications is to do a series of pairwise invader-resident comparisons, each with an 
equal weighting of invader and resident $r$ values; for our context this amounts to setting scaling factors to $1$. 
We consider both approaches because the above-mentioned shortcomings do not apply to our case, but it seems 
possible that future applications will often set scaling factors to $1$, and we want our ideas to also be considered 
by researchers making that choice.

We begin by decomposing 
the invasion growth rate,
\begin{equation}
\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})].
\end{equation}
The invasion growth rate with both $E_i$ and $C_{i \bs i}$ constant at their means is 
\begin{align}
\epsilon_i^0 = r_i(\overline{E}_i,\overline{C}_{i \bs i}).
\end{align}
The additional contribution of variation in $E_i$ to the invasion growth rate is then
\begin{equation}
\olep_i^E = \E[r_i(E_i,\overline{C}_{i \bs i})]-\epsilon_i^0,
\end{equation}
whereas the additional contribution of variation in $C_{i \bs i}$ is 
\begin{equation}
\olep_i^C=\E[r_i(\overline{E}_i,C_{i \bs i})]-\epsilon_i^0.
\end{equation}
We then define
\begin{align}
\olep_i^{EC} &= \E[r_i(E_i,C_{i \bs i})]-\olep_i^E-\olep_i^C-\epsilon_i^0 \\
&= \olr_{i \bs i} - \olep_i^E - \olep_i^C - \epsilon_i^0.
\end{align}
Thus we have the primary decomposition 
\begin{equation}
\olr_{i \bs i} = \epsilon_i^0 + \olep_i^E + \olep_i^C + \olep_i^{EC}.
\end{equation}
But some of $\olep_i^{EC}$ comes from the fact that $E_i$ and $C_{i \bs i}$ are
both varying, and some comes from covariation in these two quantities. So we next separate these
contributions in a secondary decomposition of $\olep_i^{EC}$.

As in the main text, we can now define the tertiary decomposition
\begin{align}
\olep_i^{(EC)} &= \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\shp,C_{i \bs i}^\shp)] \\
&= \left[ \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\nat,C_{i \bs i}^\nat)] \right]+\left[ \E[r_i(E_i^\nat,C_{i \bs i}^\nat)]-\E[r_i(E_i^\shp,C_{i \bs i}^\shp)] \right] \\
&= \olep_i^{[EC]}+\olep_i^{[E \nat C]}.
\end{align}
The brackets in the superscripts of $\olep_i^{[EC]}$ and $\olep_i^{[E \nat C]}$ indicate that we are talking about a term in the
tertiary decomposition, i.e., in the decomposition of $\olep_i^{(EC)}$. 
The term $\olep_i^{[EC]}$ quantifies the influence of asymmetric tail associations of $E_i$ and $C_{i \bs i}$ on the invasion growth rate,
and the the term $\olep_i^{[E \nat C]}$ corresponds to $EC$ correlations \emph{per se}, after the removal of asymmetries of tail association.
The symbol $\nat$ denotes the removal of asymmetries of tail association. We therefore have 
\begin{equation}
\olr_{i \bs i} = \epsilon_i^0 + \olep_i^E + \olep_i^C + \olep_i^{(E\shp C)}+\olep_i^{[E \nat C]}+\olep_i^{[EC]},\label{eq:decomp_overall_IGR}
\end{equation}
our overall decompsition of $r_{i \bs i}$.
The same derivation also applies to provide 
\begin{equation}
\olr_{j \bs i} = \epsilon_j^0 + \olep_j^E + \olep_j^C + \olep_j^{(E\shp C)}+\olep_j^{[E \nat C]}+\olep_j^{[EC]},\label{eq:decomp_overall_rjbsi}
\end{equation}
an analogous decomposition of $\olr_{j \bs i}$.

>>>>

%***DAN: I decided to present the general theory (not for the lottery model but for the
%general case) in the Theory section in the main text, putting off the specific lottery
%model application to later. For that, I need to explain scaling factors in general.
%So this section is now for that explanation. That means the below text, which is about
%the scaling factors for the lottery model need to be handled later. It is good, however,
%that the below text is for the general lottery model, for either way of doing the 
%fecundities. That will come in handy later.
%
% For the lottery model, for $i \neq j$, we have
% \begin{align}
% \sC_j &= -\ln \left[ 1-\delta+\delta N \frac{\overline{B}_j}{B_1(t) N_1 (t) + B_2(t) N_2(t)}  \right]  \\
% \sC_{j \bs i} &= -\ln \left[ 1-\delta+\delta  \frac{\overline{B}_j}{B_j(t)} \right] \\
% \sC_{i \bs i} &= -\ln \left[ 1-\delta+\delta  \frac{\overline{B}_i}{B_j(t)} \right].
% \end{align}
% Both $\sC_{i \bs i}$ and $\sC_{j \bs i}$ are strictly monotonic functions of $B_j$, so the function
% $B_j \mapsto \sC_{j \bs i}$ is invertible, and using its inverse provides an expression of $\sC_{i \bs i}$ 
% in terms of $\sC_{j \bs i}$. One could write this expression explicitly and then differentiate it to calculate $q_{ij}$,
% but we instead compute $q_{ij}$ through implicit differentiation:
% \begin{align}
% \frac{\partial \sC_{i \bs i}}{\partial \sC_{j \bs i}} &= \left. \frac{\partial \sC_{i \bs i}}{\partial B_j} \middle/ \frac{\partial \sC_{j \bs i}}{\partial B_j} \right. \\
% &= \left( \frac{1-\delta +\delta \frac{\overline{B}_j}{B_j}}{1-\delta+\delta \frac{\overline{B}_i}{B_j}} \right) \left( \frac{\overline{B}_i}{\overline{B}_j}  \right).
% \end{align}
% Solving $\sC_{j \bs i}=0$ gives $B_j=\overline{B}_j$, so 
% \begin{align}
% q_{ij} &= \left( \frac{1-\delta +\delta \frac{\overline{B}_j}{\overline{B}_j}}{1-\delta+\delta \frac{\overline{B}_i}{\overline{B}_j}} \right) \left( \frac{\overline{B}_i}{\overline{B}_j}  \right) \\
% &= \frac{\overline{B}_i}{\overline{B}_j (1-\delta) +\overline{B}_i \delta},
% \end{align}
% as stated in the main text.

We also use $q_{ij}=1$ as an alternative value, for the following reasons.
First, as discussed in the supporting information of \cite{Ellner_2016}, and as further elaborated by 
\cite{Barabas_2018}, it takes very special circumstances for the scaling factors to exist and be unique, 
especially with size-structured models. Second, in empirically parameterized models, scaling factor values are 
often not robust to arbitrary modelling choices. 
Although the lottery model to which we apply our theory 
is not size structured and we do not empirically parameterize it, the above shortcomings of 
scaling factors, generally, suggest we should find alternatives to their use. The alternative 
suggested by Ellner for empirical applications is to do a series of pairwise invader-resident comparisons, each with an 
equal weighting of invader and resident $r$ values; for our context this amounts to setting scaling factors to $1$. 
We consider both approaches because the above-mentioned shortcomings do not apply to our case, but it seems 
possible that future applications will often set scaling factors to $1$, and we want our ideas to also be considered 
by researchers making that choice.

We begin by decomposing the GWR,
\begin{equation}
\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})].
\end{equation}
Replacing both $E_i$ and $C_{i \bs i}$ with their means gives 
\begin{align}
\epsilon_i^0 = r_i(\overline{E}_i,\overline{C}_{i \bs i}).
\end{align}
The additional contribution of variation in $E_i$ to the GWR is then
\begin{equation}
\olep_i^E = \E[r_i(E_i,\overline{C}_{i \bs i})]-\epsilon_i^0,
\end{equation}
whereas the additional contribution of variation in $C_{i \bs i}$ is 
\begin{equation}
\olep_i^C=\E[r_i(\overline{E}_i,C_{i \bs i})]-\epsilon_i^0.
\end{equation}
We then define
\begin{align}
\olep_i^{EC} &= \E[r_i(E_i,C_{i \bs i})]-\olep_i^E-\olep_i^C-\epsilon_i^0 \\
&= \olr_{i \bs i} - \olep_i^E - \olep_i^C - \epsilon_i^0.
\end{align}
Thus we have the primary decomposition 
\begin{equation}
\olr_{i \bs i} = \epsilon_i^0 + \olep_i^E + \olep_i^C + \olep_i^{EC}.
\end{equation}
But some of $\olep_i^{EC}$ comes from the fact that $E_i$ and $C_{i \bs i}$ are
both varying, and some comes from covariation in these two quantities. So we next separate these
contributions in a secondary decomposition of $\olep_i^{EC}$.

As in the main text, we can now define the tertiary decomposition
\begin{align}
\olep_i^{(EC)} &= \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\shp,C_{i \bs i}^\shp)] \\
&= \left[ \E[r_i(E_i,C_{i \bs i})]-\E[r_i(E_i^\nat,C_{i \bs i}^\nat)] \right]+\left[ \E[r_i(E_i^\nat,C_{i \bs i}^\nat)]-\E[r_i(E_i^\shp,C_{i \bs i}^\shp)] \right] \\
&= \olep_i^{[EC]}+\olep_i^{[E \nat C]}.
\end{align}
The brackets in the superscripts of $\olep_i^{[EC]}$ and $\olep_i^{[E \nat C]}$ indicate that we are talking about a term in the
tertiary decomposition, i.e., in the decomposition of $\olep_i^{(EC)}$. 
The term $\olep_i^{[EC]}$ quantifies the influence of asymmetric tail associations of $E_i$ and $C_{i \bs i}$ on GWR,
and the the term $\olep_i^{[E \nat C]}$ corresponds to $EC$ correlations \emph{per se}, after the removal of asymmetries of tail association.
The symbol $\nat$ denotes the removal of asymmetries of tail association. We therefore have 
\begin{equation}
\olr_{i \bs i} = \epsilon_i^0 + \olep_i^E + \olep_i^C + \olep_i^{(E\shp C)}+\olep_i^{[E \nat C]}+\olep_i^{[EC]},\label{eq:decomp_overall_IGR}
\end{equation}
our overall decompsition of $r_{i \bs i}$.
The same derivation also applies to provide 
\begin{equation}
\olr_{j \bs i} = \epsilon_j^0 + \olep_j^E + \olep_j^C + \olep_j^{(E\shp C)}+\olep_j^{[E \nat C]}+\olep_j^{[EC]},\label{eq:decomp_overall_rjbsi}
\end{equation}
an analogous decomposition of $\olr_{j \bs i}$.

@@
